% BEGIN APPENDIX 2 - STUART
% Last updated by Milan Ilnyckyj 2013-06-14



		\singlespacing
		\section {Appendix II: The 200 companies with the largest fossil fuel reserves}
		\label{200Companies}
		\doublespacing



Toronto350.org is asking the University of Toronto to divest from direct stock holdings the following 200 companies, as listed in the Carbon Tracker Initiative's 2012 report, \emph{Unburnable Carbon}.\footcite{CTI2012} 
These are the top 200 listed companies ranked by estimated carbon reserves.



\begin{itemize}
  \item African Rainbow Minerals Ltd.
  \item AGL Energy
  \item Alcoa Inc.
  \item Allete Inc.
  \item Alliance Resource Partners L.P.
  \item Alpha Natural Resources Inc.
  \item Anadarko Petroleum Corp.
  \item Anglo American PLC
  \item Apache Corp.
  \item Aquila Resources Ltd.
  \item Arc Resources Ltd.
  \item ArcelorMittal
  \item Arch Coal Inc.
  \item Aston Resources Pty Ltd.
  \item ATP Oil \& Gas Corp.
  \item Bandanna Energy Ltd.
  \item Bankers Petroleum Ltd.
  \item Banpu PCL
  \item Bashneft
  \item Baytex Energy Corp.
  \item Berry Petroleum Co. (Cl A)
  \item BG Group PLC
  \item BHP Billiton
  \item Black Hills Corp.
  \item Bonavista Energy Corp
  \item BP PLC
  \item Bumi Resources
  \item Cairn Energy PLC
  \item Canadian Natural Resources Ltd.
  \item Canadian Oil Sands Ltd.
  \item Capital Power Corp.
  \item Cenovus Energy Inc.
  \item Chesapeake Energy Corp.
  \item Chevron Corp.
  \item China Shenhua Energy Co. Ltd.
  \item Churchill Mining PLC
  \item Cimarex Energy Co.
  \item Cliffs Natural Resources Inc.
  \item Cloud Peak Energy Inc.
  \item CLP Holdings Ltd.
  \item CNOOC Ltd.
  \item Coal India Ltd.
  \item Coal of Africa Ltd.
  \item Compania Espanola de Petroleos S.A.
  \item Concho Resources Inc.
  \item ConocoPhillips
  \item Consol Energy Inc.
  \item Continental Resources Inc. Oklahoma
  \item Crescent Point Energy Corp.
  \item Datang International Power Generation Co. Ltd.
  \item Datong Coal Industry Co. Ltd.
  \item Denbury Resources Inc.
  \item Devon Energy Corp.
  \item Ecopetrol S.A.
  \item El Paso Corp.
  \item EnCana Corp.
  \item Energen Corp.
  \item Enerplus Corp.
  \item ENI S.p.A.
  \item EOG Resources Inc.
  \item EQT Corp.
  \item Eurasian Natural Resources Corp. PLC
  \item Evraz Group S.A.
  \item Exxaro Resources Ltd.
  \item Exxon Mobil Corp.
  \item FirstEnergy Corp.
  \item Forest Oil Corp.
  \item Fortune Minerals Ltd.
  \item Fushan International Energy Group Ltd.
  \item Gansu Jingyuan Coal Industry \& Electricity Power 
  \item Gazprom OAO
  \item GDF Suez S.A.
  \item Global Energy Development PLC
  \item Grupo Mexico S.A.B. de C.V.
  \item Gujarat NRE Coke Ltd.
  \item Gujarat NRE Coking Coal Ltd.
  \item Hess Corp.
  \item Homeland Energy Group Ltd.
  \item Huolinhe Opencut Coal Industry Corp. Ltd.
  \item Husky Energy Inc.
  \item Idemitsu Kosan Co. Ltd.
  \item Imperial Oil Ltd.
  \item INA-Industrija Nafte
  \item Inner Mongolia Yitai Coal Co. Ltd.
  \item Inpex Corp.
  \item International Coal Group Inc.
  \item Irkutskenergo
  \item Itochu Corp.
  \item James River Coal Co.
  \item Jindal Steel \& Power Ltd.
  \item Jizhong Energy Resources Co. Ltd.
  \item Kazakhmys PLC
  \item Kuzbassenergo
  \item Linn Energy LLC
  \item Lukoil Holdings
  \item Lundin Petroleum AB
  \item Macarthur Coal Pty Ltd.
  \item Magnitogorsk Iron \& Steel Works
  \item Marathon Oil Corp.
  \item Mariner Energy
  \item Massey Energy Co.
  \item Mechel OAO
  \item Mitsubishi Corp.
  \item Mitsui \& Co. Ltd.
  \item Mitsui Matsushima Co. Ltd.
  \item MOL Hungarian Oil and Gas Plc
  \item Mongolian Mining Corp.
  \item Murphy Oil Corp.
  \item NACCO Industries Inc. (Cl A)
  \item New Hope Corp. Ltd.
  \item New World Resources N.V.
  \item Newfield Exploration Co.
  \item Nexen Inc.
  \item Neyveli Lignite Corp. Ltd.
  \item Noble Energy Inc.
  \item Noble Group Ltd
  \item Northern Energy Corp. Ltd.
  \item Novatek
  \item Novolipetsk Steel OJSC
  \item NTPC Ltd.
  \item Occidental Petroleum Corp.
  \item Oil \& Natural Gas Corp. Ltd.
  \item Oil India Ltd.
  \item Oil Search Ltd.
  \item OMV AG
  \item Optimum Coal Holdings Ltd.
  \item PA Resources AB
  \item Pacific Rubiales Energy Corp.
  \item Patriot Coal Corp.
  \item Peabody Energy Corp.
  \item Pengrowth Energy Corp.
  \item Penn West Petroleum Ltd.
  \item PetroBakken Energy Ltd.
  \item Petrobank Energy \& Resources Ltd.
  \item Petrobras
  \item Petroleum Development Corp.
  \item Pingdingshan Tianan Coal Mining Co. Ltd.
  \item Pioneer Natural Resources Co.
  \item Plains Exploration \& Production Co.
  \item Polo Resources Ltd.
  \item Polyus Gold OAO
  \item Premier Oil PLC
  \item Prophecy Resource Corp.
  \item PT Adaro Energy
  \item PT Bayan Resources
  \item PTT PCL
  \item Public Power Corp. S.A.
  \item Questar Corp.
  \item Quicksilver Resources Inc.
  \item Range Resources Corp.
  \item Raspadskaya OJSC
  \item Repsol YPF S.A.
  \item Resolute Energy Corp.
  \item Rio Tinto
  \item Rosneft
  \item Royal Dutch Shell PLC
  \item RWE AG
  \item SandRidge Energy Inc.
  \item Santos Ltd.
  \item Sasol Ltd.
  \item Severstal JSC
  \item Shanxi Coking Co. Ltd.
  \item Sherritt International Corp.
  \item SINOPEC Shandong Taishan Petroleum Co. Ltd.
  \item SK Holdings Co. Ltd.
  \item SM Energy Co.
  \item Soco International PLC
  \item Southwestern Energy Co.
  \item Statoil ASA
  \item Straits Asia Resources Ltd.
  \item Suncor Energy Inc.
  \item Swift Energy Co.
  \item Talisman Energy Inc.
  \item Tata Power Co. Ltd.
  \item Tata Steel Ltd.
  \item Teck Resources Ltd.
  \item Tokyo Electric Power Co. Inc.
  \item Total S.A.
  \item TransAlta Corp.
  \item Tullow Oil PLC
  \item Ultra Petroleum Corp.
  \item United Co. Rusal PLC
  \item United Industrial Corp. Ltd.
  \item Vale SA
  \item Venoco Inc.
  \item Walter Energy, Inc.
  \item Wescoal Holdings Ltd.
  \item Wesfarmers Ltd.
  \item Western Coal Corp.
  \item Westmoreland Coal Co.
  \item Whitehaven Coal Ltd.
  \item Whiting Petroleum Corp.
  \item Williams Cos.
  \item Woodside Petroleum Ltd.
  \item Xstrata PLC
  \item Yanzhou Coal Mining Co. Ltd.
  \item YPF S.A.
  \item Zhaikmunai L.P.
  \item Zhengzhou Coal Industry \& Electric Power Co. Ltd.
\end{itemize}



% END APPENDIX 2 - STUART