% BEGIN SECTION 1 - MILAN
% Last updated by Milan Ilnyckyj 2013-MAY-27

		\section{Executive summary}

\begin{vcom}
This will be filled in once we are reasonably happy with the rest of the brief. It will certainly be the section that the most people read and which gets the most attention, so we should make a special effort to make the case compelling here.
\end{vcom}


\begin{vcom}
	Citations (and probably internal links to later parts of the brief) should be added to this section.
\end{vcom}


The governments of the world, including the government of Canada, have agreed that raising global temperatures to more than 2˚C above where they were before we started burning fossil fuels would be ``dangerous''.\footcite[][]{CopenhagenAccord}
If we are to achieve that objective, we cannot burn most of the fossil fuels that remain on the planet.
Based on hundreds of thousands of years of evidence on how the climate responds to greenhouse gasses (GHGs), we can calculate that avoiding a 2˚C increase means we must keep future GHG pollution to no more than 565 billion tonnes (gigatonnes) of carbon dioxide (\ce{CO2}).\footcite[For an excellent summary that is accessible to non-experts see: ][]{TerrifyingNewMath}
At the same time, we know that burning the world's proven reserves of coal, oil, and natural gas would produce 2,795 gigatonnes of \ce{CO2} --- nearly five times as much as it would be safe to burn.\footcite[][]{CTI2012}
That means we need to find a way to keep 80\% of the world's fossil fuel reserves unburned.\footcite[Another accessible summary of the issue can be found in: ][]{HotBackyard}



Climate change is a defining example of social injury.
Firms that produce fossil fuels do not bear any economic burden as a result of the many forms of harm they are imposing on other people, including agricultural impacts, sea level rise, human health impacts, and more severe extreme weather.
Likewise, those who use fossil fuels enjoy the benefits while imposing these costs on others.
In order to avoid severe global injury, the total quantity of fossil fuels burned by humanity must be capped at a level far below the level of fossil fuels available to be burned.



The business plans of fossil fuel companies do not take this objective into account.
They assume they can burn all of their proven reserves, along with any additional reserves they discover in unconventional areas like the arctic, the deep ocean, and Canada's bituminous sands.
Right now, we are adding about 30 gigatonnes of \ce{CO2} to the atmosphere each year, and the amount we add is increasing at a rate of about 3\%.
That means that we are on track to exceed the 565 gigatonne limit within 15 years.



Two big implications arise from this. 
First --- we need to find a way to meet the world's energy needs without burning most of the Earth's remaining fossil fuels. 
This requires a massive redirection of investment from fossil fuel energy sources to different energy sources that do not alter the climate.
Second --- the stockmarket value of fossil fuel companies is based on the assumption that they will be able to dig up and burn what they own.
If they are allowed to do this, the global effects will be catastrophic.
As such, much of the value of these companies is an illusion, based on an out-dated assumption that we can use the atmosphere forever as a free dumping ground for \ce{CO2}.



The University of Toronto has an opportunity to do two things: to take part in the redirection of investment that is necessary to prevent climatic catastrophe, and to sell its shares in fossil fuel companies before the general public accepts that most of their reserves are unburnable.
This brief explains in detail why divestment from fossil fuel companies is in keeping with the values of the university and why it is feasible and financially prudent.



By making the choice to divest, the University of Toronto can improve its future financial prospects, uphold its values, and take part in a necessary global transition away from \ce{CO2}-intensive forms of energy.
The University of Toronto's Statement of Institutional Purpose includes ``a resolute commitment to the principles of equal opportunity, equity and justice.'' \footcite{InstitutionalPurpose}
If future generations are to have equal opportunities, they cannot inherit a planet that has been impoverished by uncontrolled climate change.
Similarly, the principles of equity and justice forbid us from ignoring what we know about the harms of greenhouse gas pollution by continuing to impose risk and suffering on innocent people around the world and in future generations.



Across North America, the peer schools of the University of Toronto are considering divestment.
There include Harvard, Yale, Princeton, Stanford, and MIT.
[NUMBER] schools have already committed to divest.
By leading the way and divesting first, the University of Toronto can distinguish itself as being ahead of the pack on one of the major issues of the 21st century.
There is also a convincing argument that those who have benefitted substantially from the use of fossil fuels to date have a particular obligation to help the world transition onto a low-carbon development pathway.
The per-capita emissions of Canadians are extremely high, and both Canadians in general and the University of Toronto in particular occupy privileged places in the world.
By showing moral leadership, the University of Toronto can demonstrate the degree to which it understands the major trends now shaping the world, and its determination to end up on the right side of history, both in terms of upholding values and in terms of protecting its financial security.


% END SECTION 1 - MILAN